---
title: "Mixed ANOVA: anxiety across treatment groups and timepoints"
subtitle: "Clinical trial: CBT vs pharmacological vs control"
author: "Tu nombre"
date: today
date-format: "D MMMM YYYY"
format:
  html:
    theme: cosmo
    toc: true
    toc-location: left
    embed-resources: true
    code-fold: true              # codigo oculto por defecto, expandible al click
    code-tools: false
    grid:
      body-width: 900px
      sidebar-width: 250px
      margin-width: 200px
execute:
  echo: true
  eval: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(afex)
library(emmeans)
library(effectsize)
library(knitr)
library(jtools)

options(scipen = 999)

data_processed <- read_csv("../exercises/data/processed_data/preprocessed_data.csv")

data_long <- data_processed |>
  pivot_longer(
    cols          = c(starts_with("anxiety"), starts_with("depression"),
                      starts_with("wellbeing"), starts_with("loneliness"),
                      starts_with("self_esteem")),
    names_pattern = "(.*)_(pre|post|followup)",
    names_to      = c(".value", "timepoint")
  ) |>
  mutate(timepoint = factor(timepoint, levels = c("pre", "post", "followup")))
```

# Methods

A randomized clinical trial assigned 300 patients to one of three conditions: cognitive-behavioral therapy (CBT, *n* = 100), pharmacological treatment (*n* = 100), or waitlist control (*n* = 100). Anxiety was measured with the GAD-7 at three timepoints: pre-treatment, post-treatment, and 3-month follow-up.

A 3 (group: CBT vs. pharmacological vs. control) × 3 (timepoint: pre vs. post vs. follow-up) mixed ANOVA was conducted on GAD-7 scores using the `afex` R package, with Greenhouse–Geisser correction for sphericity. Post-hoc pairwise comparisons used Holm correction. Effect sizes are reported as generalized eta squared ($\eta^2_G$; Olejnik & Algina, 2003).

# Results

## Descriptive statistics

```{r}
#| label: descriptives

descriptivos <- data_long |>
  summarise(
    M  = round(mean(anxiety), 2),
    SD = round(sd(anxiety), 2),
    n  = n(),
    .by = c(group, timepoint)
  ) |>
  arrange(group, timepoint)

descriptivos |>
  kable(
    format    = "html",
    caption   = "Table 1. Descriptive statistics for GAD-7 anxiety scores by group and timepoint.",
    col.names = c("Group", "Timepoint", "M", "SD", "n")
  )
```

## Mixed ANOVA

```{r}
#| label: model

modelo_mixto <- aov_ez(
  id      = "patient_id",
  dv      = "anxiety",
  data    = data_long,
  within  = "timepoint",
  between = "group"
)

afex::nice(modelo_mixto) |>
  kable(
    format    = "html",
    caption   = "Table 2. Mixed ANOVA results for GAD-7 anxiety: timepoint x group.",
    col.names = c("Effect", "df", "MSE", "F", "ges", "p"),
    escape    = FALSE
  )
```

The mixed ANOVA revealed significant main effects of timepoint and group, as well as a significant timepoint × group interaction (see Table 2). The interaction effect was of primary interest: it indicates that the change in anxiety over time **differed across treatment conditions**.

## Effect sizes with confidence intervals

```{r}
#| label: effect-size

eta_squared(model = modelo_mixto, generalized = TRUE) |>
  kable(
    format  = "html",
    caption = "Table 3. Generalized eta squared with 95% confidence intervals.",
    digits  = 3
  )
```

## Visualization

```{r}
#| label: fig-trayectorias
#| fig-cap: "Figure 1. Mean GAD-7 anxiety across timepoints by treatment group. Error bars represent ±1 standard error of the mean."
#| fig-width: 9
#| fig-height: 5

fig_trayectorias_anxiety <- ggplot(
  data    = descriptivos,
  mapping = aes(x = timepoint, y = M, color = group, group = group)
) +
  geom_point(size = 3) +
  geom_line(linewidth = 1) +
  geom_errorbar(
    aes(ymin = M - SD/sqrt(n), ymax = M + SD/sqrt(n)),
    width = 0.1
  ) +
  labs(
    x     = "Timepoint",
    y     = "GAD-7 anxiety (mean)",
    color = "Group"
  ) +
  theme_apa()

fig_trayectorias_anxiety

ggsave(
  filename = "../exercises/output/fig_trayectorias_anxiety.png",
  plot     = fig_trayectorias_anxiety,
  width    = 9, height = 5, dpi = 300
)
```

```{r}
#| label: fig-interaction
#| fig-cap: "Figure 2. Estimated marginal means from the mixed ANOVA model (95% CI)."
#| fig-width: 9
#| fig-height: 5

fig_interaccion_anxiety <- emmip(
  object  = modelo_mixto,
  formula = group ~ timepoint,
  CIs     = TRUE
) +
  labs(
    x = "Timepoint",
    y = "GAD-7 anxiety (estimated mean)"
  ) +
  theme_apa()

fig_interaccion_anxiety

ggsave(
  filename = "../exercises/output/fig_interaccion_anxiety.png",
  plot     = fig_interaccion_anxiety,
  width    = 9, height = 5, dpi = 300
)
```

## Post-hoc comparisons

### Time effects within each group

```{r}
#| label: posthoc-within

emm_within <- emmeans(object = modelo_mixto, specs = ~ timepoint | group)

pairs(x = emm_within, adjust = "holm") |>
  as.data.frame() |>
  kable(
    format  = "html",
    caption = "Table 4. Pairwise comparisons of timepoints within each group (Holm-corrected).",
    digits  = 3
  )
```

### Group differences at each timepoint

```{r}
#| label: posthoc-between

emm_between <- emmeans(object = modelo_mixto, specs = ~ group | timepoint)

pairs(x = emm_between, adjust = "holm") |>
  as.data.frame() |>
  kable(
    format  = "html",
    caption = "Table 5. Pairwise comparisons of groups at each timepoint (Holm-corrected).",
    digits  = 3
  )
```

# Interpretation

The significant `timepoint × group` interaction confirms that the three treatment conditions produced **different patterns of change in anxiety over time**. Post-hoc comparisons (Tables 4 and 5) clarify the pattern:

- **CBT** showed a substantial reduction in anxiety from pre to post and **maintained** the reduction at follow-up.
- **Pharmacological treatment** produced a comparable reduction at post-treatment but showed a **partial rebound** at follow-up, consistent with treatment discontinuation effects.
- **Control** participants showed minimal change across timepoints.

These findings are consistent with the literature suggesting that the benefits of CBT for anxiety disorders tend to be **maintained beyond the end of active treatment**, whereas pharmacological effects may attenuate after discontinuation.

# References

Cohen, J. (1988). *Statistical power analysis for the behavioral sciences* (2nd ed.). Lawrence Erlbaum Associates.

Olejnik, S., & Algina, J. (2003). Generalized eta and omega squared statistics: Measures of effect size for some common research designs. *Psychological Methods*, *8*(4), 434–447.

Singmann, H., Bolker, B., Westfall, J., Aust, F., & Ben-Shachar, M. S. (2024). *afex: Analysis of factorial experiments* [R package]. https://CRAN.R-project.org/package=afex

# Session information

```{r}
#| label: session-info
#| code-fold: true
sessionInfo()
```
